---
title: "4: Part 1 - Data Wrangling"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the usefulness of data wrangling and its place in the data pipeline
2. Wrangle datasets with dplyr functions
3. Apply data wrangling skills to a real-world example dataset

## Set up your session

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 

Opening discussion: why might we be interested in long-term observations of temperature, oxygen, and light in lakes?

> Add notes here: 

```{r, message = FALSE}
getwd()
#install.packages("tidyverse")
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
NTL.phys.data <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

colnames(NTL.phys.data)
head(NTL.phys.data)
summary(NTL.phys.data)
str(NTL.phys.data)
dim(NTL.phys.data)

class(NTL.phys.data$sampledate)
# Format sampledate as date
NTL.phys.data$sampledate <- as.Date(NTL.phys.data$sampledate, format = "%m/%d/%y")
```

## Data Wrangling

Data wrangling extends data exploration: it allows you to process data in ways that are useful for you. An important part of data wrangling is creating *tidy datasets*, with the following rules: 

1. Each variable has its own column
2. Each observation has its own row
3. Each value has its own cell

What is the best way to wrangle data? There are multiple ways to arrive at a specific outcome in R, and we will illustrate some of those approaches. Your goal should be to write the simplest code that will get you to your desired outcome. However, there is sometimes a trade-off of the opportunity cost to learn a new formulation of code and the time it takes to write complex code that you already know. Remember that the best code is one that is easy to understand for yourself and your collaborators. Remember to comment your code, use informative names for variables and functions, and use reproducible methods to arrive at your output.

## Dplyr Wrangling Functions

`dplyr` is a package in R that includes functions for data manipulation (i.e., data wrangling or data munging). `dplyr` is included in the tidyverse package, so you should already have it installed on your machine. The functions act as verbs for data wrangling processes. For more information, run this line of code:

```{r, results = "hide"}
vignette("dplyr")
```

### Filter

Filtering allows us to choose certain rows (observations) in our dataset.

#choose rows based on column values

Here are the relevant commands used in the `filter` function. Add some notes to designate what these commands mean. 
`==`
`!=`
`<`
`<=`
`>`
`>=`
`&`
`|`

```{r}
class(NTL.phys.data$lakeid)
class(NTL.phys.data$depth)

# matrix filtering
#we want to look at surface data on the lake
#filtering rows that have a zero on column depth
NTL.phys.data.surface1 <- NTL.phys.data[NTL.phys.data$depth == 0,]

# dplyr filtering (another method)
NTL.phys.data.surface2 <- filter(NTL.phys.data, depth == 0)
NTL.phys.data.surface3 <- filter(NTL.phys.data, depth < 0.25)

# Did the methods arrive at the same result?
head(NTL.phys.data.surface1)
dim(NTL.phys.data.surface1)
head(NTL.phys.data.surface2)
dim(NTL.phys.data.surface2)
head(NTL.phys.data.surface3)
dim(NTL.phys.data.surface3)

# Choose multiple conditions to filter
summary(NTL.phys.data$lakename)
#Selecting data associated only with Peter Lake and Paul Lake

#Selecting all instances of Peter OR Paul lake in the code below
NTL.phys.data.PeterPaul1 <- filter(NTL.phys.data, lakename == "Paul Lake" | lakename == "Peter Lake")
#Doing the same thing that the code above does but in a different way
NTL.phys.data.PeterPaul2 <- filter(NTL.phys.data, lakename != "Central Long Lake" & 
                                     lakename != "Crampton Lake" & lakename != "East Long Lake" &
                                     lakename != "Hummingbird Lake" & lakename != "Tuesday Lake" &
                                     lakename != "Ward Lake" & lakename != "West Long Lake")
# %in% is a cleaner way of filtering data to select data only associated with Peter and Paul Lake
NTL.phys.data.PeterPaul3 <- filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake"))

# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
#looking for range from the first day of June to the last day of October
NTL.phys.data.JunethruOctober1 <- filter(NTL.phys.data, daynum > 151 & daynum < 305)
#same thing as code above but with a comma instead of &
NTL.phys.data.JunethruOctober2 <- filter(NTL.phys.data, daynum > 151, daynum < 305)
NTL.phys.data.JunethruOctober3 <- filter(NTL.phys.data, daynum >= 152 & daynum <= 304)
NTL.phys.data.JunethruOctober4 <- filter(NTL.phys.data, daynum %in% c(152:304))
#all of the code above does the same thing but in a different way with the filter function

# Exercise: 
# filter NTL.phys.data for the year 1999
NTL.phys.data.1999_data <- filter(NTL.phys.data, year4 > 1998, year4 <2000)
# what code do you need to use, based on the class of the variable?
class(NTL.phys.data$year4)
#code says that year4 is classified as an integer

# Exercise: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.

NTL.phys.data.TuesdayLake <- filter(NTL.phys.data, lakename %in% c("Tuesday Lake"))

NTL.phys.data.TuesdayLake_1990_to_1999 <- filter(NTL.phys.data.TuesdayLake, year4 > 1989, year4 < 2000)

```
Question: Why don't we filter using row numbers?

> Answer: 

### Arrange

Arranging allows us to change the order of rows in our dataset. By default, the arrange function will arrange rows in ascending order.

```{r}
#change order of a specified row
NTL.phys.data.depth.ascending <- arrange(NTL.phys.data, depth)
view(NTL.phys.data)
NTL.phys.data.depth.descending <- arrange(NTL.phys.data, desc(depth))

# Exercise: 
# Arrange NTL.phys.data by temperature, in descending order. 
# Which dates, lakes, and depths have the highest temperatures?

NTL.phys.data.temperature.ascending <- arrange(NTL.phys.data, desc(temperature_C))
# Depths: ~0 - 0.5 m
# Lakes: East Long Lake and Hummingbird Lake
# Dates: 7/16/1998, 7/1/2002, 7/19/2002 ...
```
### Select

Selecting allows us to choose certain columns (variables) in our dataset.

```{r}
#Want columns sampled ate to temperature_C in the code below
NTL.phys.data.temps <- select(NTL.phys.data, lakename, sampledate:temperature_C)

```
### Mutate

Mutating allows us to add new columns that are functions of existing columns. Operations include addition, subtraction, multiplication, division, log, and other functions.

```{r}
#Converting temp. in C to temmp in F
NTL.phys.data.temps <- mutate(NTL.phys.data.temps, temperature_F = (temperature_C*9/5) + 32)

```

## Lubridate

A package that makes coercing date much easier is `lubridate`. A guide to the package can be found at https://lubridate.tidyverse.org/. The cheat sheet within that web page is excellent too. This package can do many things (hint: look into this package if you are having unique date-type issues), but today we will be using two of its functions for our NTL dataset. 

```{r}
# add a month column to the dataset
NTL.phys.data.PeterPaul1 <- mutate(NTL.phys.data.PeterPaul1, month = month(sampledate)) 
view(NTL.phys.data.PeterPaul1)
# reorder columns to put month with the rest of the date variables
NTL.phys.data.PeterPaul1 <- select(NTL.phys.data.PeterPaul1, lakeid:daynum, month, sampledate:comments)

# find out the start and end dates of the dataset
interval(NTL.phys.data.PeterPaul1$sampledate[1], NTL.phys.data.PeterPaul1$sampledate[21613])
#use the code below if the dates are not in order 
interval(first(NTL.phys.data.PeterPaul1$sampledate), last(NTL.phys.data.PeterPaul1$sampledate))
```


## Pipes

Sometimes we will want to perform multiple functions on a single dataset on our way to creating a processed dataset. We could do this in a series of subsequent functions or create a custom function. However, there is another method to do this that looks cleaner and is easier to read. This method is called a pipe. We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r}
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)
  
view(NTL.phys.data.processed)
```

Notice that we did not place the dataset name inside the wrangling function but rather at the beginning.

### Saving processed datasets

```{r}
#creating a csv for your processed data, very usefule
write.csv(NTL.phys.data.PeterPaul1, row.names = FALSE, file = "./Data/Processed/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
```

## Closing Discussion

When we wrangle a raw dataset into a processed dataset, we create a code file that contains only the wrangling code. We then save the processed dataset as a new spreadsheet and then create a separate code file to analyze and visualize the dataset. Why do we keep the wrangling code separate from the analysis code?


