---
title: "Assignment 2: Coding Basics"
author: "Kelsey"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
#creating a sequence from 1 to 10 that increases by 4
seq(1, 100, 4)  # from, to, by
#Renaming Sequence
A02_sequence <- seq(1, 100, 4)
A02_sequence #run to check solution

#2.
#finding the mean and median of the sequence produced above
mean(A02_sequence)
median(A02_sequence)

#3. 
#determining whether the mean is greater than the median
mean(A02_sequence) > median(A02_sequence)
#determining whether the median is greater than the mean
mean(A02_sequence) > median(A02_sequence)
#determining whether the median is equal to the mean
mean(A02_sequence) == median(A02_sequence)
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5
#Create a series of vectors, each with four components: names, test scores,P/F
#vector 1; Character Vector
name_vect <- c("Natalie", "Michael", "Valerie", "Maddie")
name_vect 
#vector 2; Numeric Vector
test_score_vect <- c(55, 43, 97, 36)
test_score_vect
#vector 3; Logical Vector
pass_fail_vect <- c(TRUE, FALSE, TRUE, FALSE)
pass_fail_vect

#identifying who specifically passes the test and identifying their specific scores
test_score_vect[c(TRUE, FALSE, TRUE, FALSE)]
name_vect[c(TRUE, FALSE, TRUE, FALSE)]
#another way to see which specific scores are considers as passing
test_score_vect >= 50

#6 Completed above!

#7
#create a data frame for the information above
grade_analysis <- data.frame(Students = c("Natalie", "Michael", "Valerie", "Maddie"),
                            Test_scores = c(55, 43, 97, 36),
                            Pass_fail = c(TRUE, FALSE, TRUE, FALSE))

grade_analysis

#8
#name the columns
#Completed above!
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A data frame is more general than a matrix. In a data frame, while the list of vectors must be equal in length, the vectors can be in different modes (i.e., numbers, characters, etc.). In contrast, a matrix must contain the same type of data (i.e., in the same mode) in addition to containing vectors that are equal in length.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#Using 'if' and 'else' method
passing_grades_v1 <- function(x) {
  if(x > 50) {
    x = TRUE
  }
  else {
    x = FALSE
  }
}
Natalie_results_v1 <- passing_grades_v1(55); Natalie_results_v1
Michael_results_v1 <- passing_grades_v1(43); Michael_results_v1
Valerie_results_v1 <- passing_grades_v1(97); Valerie_results_v1
Maddie_results_v1 <- passing_grades_v1(36); Maddie_results_v1

#Using 'ifelse' statement version 1
grades <-c(55, 43, 97, 36)
ifelse(grades>50, "PASS", "FAIL") #log_exp, if TRUE, if FALSE


#Using 'ifelse' statement version 2
passing_grades_v2 <- function(x){
  ifelse(x>50, "PASS", "FAIL") #log_exp, if TRUE, if FALSE
  
}


Natalie_results_v2 <- passing_grades_v2(55); Natalie_results_v2
Michael_results_v2 <- passing_grades_v2(43); Michael_results_v2
Valerie_results_v2 <- passing_grades_v2(97); Valerie_results_v2
Maddie_results_v2 <- passing_grades_v2(36); Maddie_results_v2
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Technically both options can work but the 'ifelse' statement is far more efficient than the 'if' and 'else' version with respect to vectors. Moreover, the 'ifelse' statement is intended for vector lengths > 1. Meanwhile, the 'if' and 'else' statement is designed for items that have a length of 1.
