---
title: 'Assignment 5: Data Visualization'
author: "Kelsey Husted"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.
```{r, echo = FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE, echo=TRUE)
```


```{r}
#1 
#set working directory
getwd()
setwd("/Users/kelseyhusted 1/Desktop/Data Analytics/Environmental_Data_Analytics_2022")
#load packages
library(tidyverse)
library(cowplot)
#upload processed data
PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

NEON_NIWO_Litter <-read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2
#format date
class(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

class(NEON_NIWO_Litter$collectDate)
NEON_NIWO_Litter$collectDate <- as.Date(NEON_NIWO_Litter$collectDate, format = "%Y-%m-%d")
class(PeterPaul.chem.nutrients$month)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
my_theme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom", legend.text=element_text(size=8), legend.title=element_text(size=10), panel.background = element_rect(fill = "aliceblue"), panel.grid.major = element_line(size = 0.5, linetype = 'solid', colour = "white"))

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
TP_PO4_Plot <- ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point(alpha = 0.5, size = 1.5) +
  xlim(0, 35) +
  ylim(0, 150) +
  geom_smooth(method = lm, color = "black") +
  xlab(expression("PO"[4])) + ylab("TP ug/L") +
  scale_color_brewer(palette = "Set2") +
  my_theme
print(TP_PO4_Plot)

#A way to set the origin of plot to zero
  #scale_x_continuous(expand = c(0, 0), limits = c(0, 35)) +
  #scale_y_continuous(expand = c(0, 0), limits = c(0, 150)) +
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r, fig.height = 3, fig.width = 6}
#5
#transform month into categories
PeterPaul.chem.nutrients$month <- as.factor(PeterPaul.chem.nutrients$month)
class(PeterPaul.chem.nutrients$month)

Temp_Boxplot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  xlab("Month") + ylab("Temperature C") +
  scale_color_brewer(palette = "Set2") +
  my_theme

print(Temp_Boxplot)

TP_Boxplot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  xlab("Month") + ylab("TP ug") +
  scale_color_brewer(palette = "Set2") +
  my_theme
print(TP_Boxplot)

TN_Boxplot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  xlab("Month") + ylab("TN ug") +
  scale_color_brewer(palette = "Set2") +
  my_theme
print(TN_Boxplot)


#make legend
legend_b <- get_legend(TP_Boxplot + theme(legend.position="bottom"))
#Combine plots
Plots_Combined <- plot_grid(Temp_Boxplot + theme(legend.position = "none"), TP_Boxplot + theme(legend.position = "none"), TN_Boxplot + theme(legend.position = "none"), labels = c('A', 'B', 'C'), label_size = 12, axis = "b", align = "h", nrow = 1)

#Final figure with legend and plots
NutrientTemp_3plots <- plot_grid(Plots_Combined, legend_b, ncol = 1, rel_heights = c(1, .2))
print(NutrientTemp_3plots)


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: For the temperature plot, the later summer months (i.e., July-September) display higher temperature values in both lakes with Paul lake being slightly warmer than Peter Lake. Both TP and TN plots show a lot of variation. Even so,the nutrient Peter Lake boxplots display a higher spread and larger median during the later summer months compared to Paul Lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
NEON_NIWO_Litter_Needles <-
  ggplot(subset(NEON_NIWO_Litter, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  xlab("Collection Date") + ylab("Dry Mass of Needles") +
  scale_color_brewer(palette = "Dark2") +
  my_theme
print(NEON_NIWO_Litter_Needles)

#7
NEON_NIWO_Litter_Needles_Class <-
  ggplot(subset(NEON_NIWO_Litter, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  xlab("Collection Date") + ylab("Dry Mass of Needles") +
  scale_color_brewer(palette = "Dark2") +
  #facet_wrap(vars(nlcdClass)) +
  facet_wrap("nlcdClass", nrow = 2) +
  my_theme
  
print(NEON_NIWO_Litter_Needles_Class)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:Plot 7 is more effective because it's easier to compare the needle dry mass in each landcover type throughout the provided years without all of the landcover classes overlapping.
