---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Kelsey Husted"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
#set working directory
getwd()
setwd("/Users/kelseyhusted 1/Desktop/Data Analytics/Environmental_Data_Analytics_2022")
#load packages
library(tidyverse)
library(ggplot2)
library(agricolae)
library(lubridate)
#Import data
NTL.LTER.chem.phys <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
view(NTL.LTER.chem.phys)
#Set date
NTL.LTER.chem.phys$sampledate <- as.Date(NTL.LTER.chem.phys$sampledate, format = "%m/%d/%y")
class(NTL.LTER.chem.phys$sampledate)

#2
#Build a ggplot theme
my_theme2 <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom", legend.text=element_text(size=8), legend.title=element_text(size=10), panel.background = element_rect(fill = "aliceblue"), panel.grid.major = element_line(size = 0.5, linetype = 'solid', colour = "white"))
#set theme default
theme_set(my_theme2)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature does not change with depth (i.e., the slope and intercept are both equal to zero).
Ha: Mean lake temperature changes with depth (i.e., there is a statistically significant relationship between depth and temperature).


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
#Wrangle data
NTL.LTER.chem.phys.wrangled <-
  NTL.LTER.chem.phys %>%
  mutate(month = month (sampledate)) %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit() 
view(NTL.LTER.chem.phys.wrangled)
dim(NTL.LTER.chem.phys.wrangled)

#5
#plot continuous relationship (depth = x, temperature = y)
NTL.LTER.TempbyDepth <- 
  ggplot(subset(NTL.LTER.chem.phys.wrangled, temperature_C >= 0, temperature_C <= 35 ),
         aes(x = depth, y = temperature_C)) +
  xlim(0,16) +
  ylim(0,35) +
  labs(x = "Depth", y = "Temperature C") +
  geom_smooth(method = "lm") +
  geom_point()
print(NTL.LTER.TempbyDepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests that temperature decreases as depth increases in the lakes. The distribution of the points show that the the linearity of the trend is not perfect since the linearity plateaus around a depth of ~10m where sample points start to dwindle. Additionally, most of the points are aggregated in between 0m and 10m depths which shows that there was a greater occurance of sampling for depths closer to the surface.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
#linear regression temp by depth
temp.depth.regression <- lm(data = NTL.LTER.chem.phys.wrangled, temperature_C ~ depth)
summary(temp.depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Since the R-squared value is 0.7387, depth explains 73.87% of the variation in temperature. While the degrees of freedom is 5530, the p-value is less than 0.05 causing us to reject the null hypothesis. Furthermore, the result of the linear regression is  significant and the coefficients are statistically different than zero. The slope of the linear regression equals -1.95. Consequently, temperature decreases 1.95 C for every 1m change in depth.

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTL.LTER.AIC <- lm(data = NTL.LTER.chem.phys.wrangled, temperature_C ~ year4 + daynum + depth)
step(NTL.LTER.AIC)

#10
#all variables were recommended to predict temperature since they increase the AIC value if removed (smallest AIC is optimal).
NTL.LTER.AIC <- lm(data = NTL.LTER.chem.phys.wrangled, temperature_C ~ year4 + daynum + depth)
summary(NTL.LTER.AIC)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests to use includes the year, the day (i.e., daynum), and depth of the lakes. With a R-squared value of 0.7412, the model explains 74.12% of the observed variance which is a slight improvement compared to using solely depth as the explanatory variable which explained ~73.87% of the observed variance.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#model 1 ANOVA
NTL.LTER.anova <- aov(data = NTL.LTER.chem.phys.wrangled, temperature_C ~ lakename)
summary(NTL.LTER.anova)

#model 2 linear model
NTL.LTER.lm <- lm(data = NTL.LTER.chem.phys.wrangled, temperature_C ~ lakename)
summary(NTL.LTER.lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, according to both models, there is a significant difference between mean temperatures among lakes since the p-value is less than 0.05. Additionally, the linear model reports a R-squared value of 0.03953 which indicates that only 3.953% of the observed variance is attributed to the explanatory variable.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
TemperaturebyDepth.Plot <-
  ggplot(NTL.LTER.chem.phys.wrangled, aes(x = depth, y = temperature_C, color = lakename)) +
  ylim(0,35) +
  geom_point(alpha = 0.5) +
  labs( x = "Depth", y = "Temperature C") +
  geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "bottom")
print(TemperaturebyDepth.Plot)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
#The Anova test tells you that the means are not the same; Post-hoc test goes a step further and tells you which means are different specifically
#the null hypothesis is that the two means are equal
TukeyHSD(NTL.LTER.anova)

#group = TRUE means that you want them to actually group levels that have the same mean
NTL.LTER.anova.groups <- HSD.test(NTL.LTER.anova, "lakename", group = TRUE)
NTL.LTER.anova.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward and Paul Lake both have the same mean temperature as Peter Lake. Additionally, none of the lakes have a mean temperature that is statistically distinct from all other lakes.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample t-test could be applied to see whether Peter and Paul Lake have distinct mean temperatures. A two-sample t-test is a method used to test whether the means of two groups are equal or not.

