---
title: "Assignment 09: Data Scraping"
author: "Kelsey Husted"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
#set wd, load packages, set ggplot theme
getwd()

library(tidyverse)
library(rvest)
library(lubridate)

#theme
#set theme
my_theme1 <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom", legend.text=element_text(size=8), legend.title=element_text(size=10), panel.background = element_rect(fill = "aliceblue"), panel.grid.major = element_line(size = 0.5, linetype = 'solid', colour = "white"))
#set theme default
theme_set(my_theme1)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
#Fetch the web resources from the URL for 2020
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- webpage %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
water.system.name

pwsid <- webpage %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
pwsid

ownership <- webpage %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
ownership

max.withdrawals.mgd <- webpage %>% html_nodes("th~ td+ td") %>% html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
#creating 'out-of-order' month vector to match max-withdrawal values
month.max.withdrawals <- c("Jan", "May", "Sep", "Feb", "June", "Oct", "March", "July", "Nov", "April", "Aug", "Dec")

df <- data.frame(
  "Month" = month.max.withdrawals,
  "month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
  "Year" = rep(2020, 12),
  "Max.Withdrawals.mgd" = as.numeric(max.withdrawals.mgd),
  "Water.System.Name" = water.system.name,
  "PWSID" = pwsid,
  "Ownership" = ownership) %>%
  mutate(Date = my(paste0(month,"-",Year))) 

  
#5
#Plot the max daily withdrawals across the months for 2020
ggplot(df, aes(x = Date, y = Max.Withdrawals.mgd)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  ylab("Max Monthly Withdrawals (mgd)") +
  labs(title = "2020 Durham Water Usage Data") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels="%b, %Y", date_breaks="1 month")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(the_year, PWSID){
  
  #Retrieve website
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', PWSID, '&year=', the_year))
  
  #Scrape the data
  water.system.name <- the_website %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  pwsid <- the_website %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
  ownership <- the_website %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  max.withdrawals.mgd <- the_website %>% html_nodes("th~ td+ td") %>% html_text()
  
  #Convert to a dataframe
  df.maxwithdrawals <- tibble(
  "Month" = month.max.withdrawals,
  "month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
  "Year" = rep(the_year, 12),
  "Max.Withdrawals.mgd" = as.numeric(max.withdrawals.mgd),
  "Water.System.Name" = water.system.name,
  "PWSID" = pwsid,
  "Ownership" = ownership) %>%
  mutate(Date = my(paste0(month,"-",Year))) 
  
  #Return the dataframe
  return(df.maxwithdrawals)
}

#Run the function to test
df.functions <- scrape.it(2015, '03-32-010')
view(df.functions)

  
```


7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
view(df.functions)
ggplot(df.functions, aes(x = Date, y = Max.Withdrawals.mgd, group = 1)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE, color = "purple") +
  ylab("Max Monthly Withdrawals (mgd)") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels="%b, %Y", date_breaks="1 month") +
  labs(title = paste("2015 Water Usage Data"),
       subtitle = water.system.name)
                     
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.2015.data}
#8
#extracting 2015 Asheville info
df.functions.Asheville <- scrape.it(2015, '01-11-010')
view(df.functions.Asheville)

#Combine data
# merge two data frames by 
Combined <- rbind(df.functions.Asheville, df.functions)
view(Combined)

#plot
ggplot(Combined, aes(x = Date, y = Max.Withdrawals.mgd, color = Water.System.Name, group =Water.System.Name)) +
  ylab("Max Monthly Withdrawals (mgd)") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels="%b, %Y", date_breaks="1 month") +
  labs(title = paste("2015 Water Usage Data")) +
  legend(legend = c("Durham", "Asheville"))
  geom_line()
  
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9
#scrape years from 2010-2019
the_years = rep(2010:2019)
PWSID.Ash = '01-11-010'

#purrr's map function
the_dfs <- map(the_years,scrape.it,pwsidAsh=PWSID.Ash)

the_dfs <- lapply(X = the_years,
                  FUN = scrape.it,
                  pwsid = PWSID.Ash)



#Use lapply to apply the scrape function
the_dfs <- lapply(X = the_years,
                  FUN = scrape.it,
                  the_facility=my_facility)

#OR, use purrr's map function
the_dfs <- map(the_years,scrape.it,the_facility=my_facility)

#Conflate the returned dataframes into a single dataframe
the_df <- bind_rows(the_dfs)

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?